A Journey from Wild to Textbook Data to Reproducibly Refresh the Wages Data from the National Longitudinal Survey of Youth Database

### 2022-02-14 paper/paper.Rmd
Add explanation in dropouts section.

In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (`gender`), race (`race`), age (`age_1979`), highest grade completed (`hgc`), highest grade completed in terms of years, e.g. 9th grade = 9, 3rd year college = 15, (`hgc_i`), and the whether the graduate equivalency diploma is obtained (`ged`).

@@

<!--AE comment: I was confused by the following section: “We choose to use this revised May data because it seemed to have less missing and presumably has been checked. However, there is no revised May data for 2012, 2014, 2016, and 2018, thus, we use the ordinary May data for these years.” I was assuming the data was collected on a yearly basis, but this section suggests it was collected each month. Please clarify somewhere in the paper. 
-->
The next step is tidying to obtain `hgc` and `ged`. As the highest grade completed is already exist in  `new_data_qnames` and stored in years unit, we only rename the column name to be `hgc_i` and recode it so we have the factor data type. For `ged`, we only separate the years and the GED status of the individual. Later on, this variable is useful to determine the refreshed high school dropouts data.

@@

@@

>>>>

In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (`gender`), race (`race`), age (`age_1979`), highest grade completed (`hgc`), highest grade completed in terms of years, e.g. 9th grade = 9, 3rd year college = 15, (`hgc_i`), and whether the graduate equivalency diploma is obtained (`ged`).

@@

<!--AE comment: I was confused by the following section: “We choose to use this revised May data because it seemed to have less missing and presumably has been checked. However, there is no revised May data for 2012, 2014, 2016, and 2018, thus, we use the ordinary May data for these years.” I was assuming the data was collected on a yearly basis, but this section suggests it was collected each month. Please clarify somewhere in the paper. 
-->
The next step is tidying to obtain `hgc` and `ged`. As the highest grade completed is already exist in  `new_data_qnames` and stored in year unit, we only rename the column name to be `hgc_i` and recode it so we have the factor data type. For `ged`, we only separate the years and the GED status of the individual. Later on, this variable is useful to determine the refreshed high school dropouts data.

@@

In the original data, there is no explicit explanation of how the dropouts cohort is obtained from the database. Hence, we use the high school dropouts criteria mentioned in @nlsy79edu to determine who is the dropouts, which are:

1. The individual whose highest grade completed (`hgc`) is reported to be less than 12th grade, **or**
2. The individual whose highest grade completed (`hgc`) is reported to be at least 12th grade and have received a GED (`ged` is code to 2).

Further, @SingerJudithD2003Alda set additional criteria for the dropouts in the textbook data, which are male and aged between 14 and 17 years old in 1979. Hence, we also use this information to re-create the dropouts data.

With this filtering, we obtained 670 individuals in the refreshed data compared to 888 individuals in the original data. We then matched the individuals in the original and refreshed dataset to investigate why some individuals in the original data were not captured in the refreshed data. We found several reasons for this as follows:

1. 173 individuals were more than 17 years old in 1979. Thus, it looks like the description of the original data is not quite accurate, that there are people older than 17 in their subset. Our decision is to also include them in the refreshed data as the data has age variable in it, so analyst could filter them later. 
2. 79 individuals were less than or equal to 17 years old in 1979. However, they were not captured in the refreshed data because:

  + 35 of them completed at least 12th grade and having a diploma instead of GED (`ged` variable is coded to 1). Thus, they are indeed are not dropouts and we decide to not include them in the refreshed data. 
  + The information of `ged` is missing in 38 individuals. We decide to include them in the refreshed data. 
  + 3 individuals are having both diploma and GED (`ged` is coded to 3). They essentially have GED, so we decide to keep them in the refreshed data as well. 
  + 12 individuals do not exist in `wages` data because they were participated in less than 3 rounds of survey.

We then redo the filtering based on those decisions we made on those individuals who were not captured in the refreshed data. Finally, we have 863 dropouts in the refreshed data, the closest number of individuals we get to the original data.

@@

## The takeaways

### 2022-02-14 paper/paper.tex
Add explanation in dropouts section.

Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques. An example of this, is the wages data made public by Singer and Willett (2003) which can be used to teach generalized linear models, including hierarchical, mixed effects and multilevel. The freely available data records hourly wages of a sample of high school dropouts, from 1979-1994, along with demographic variables including education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S. Department of Labor 2021a).

The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average. An exploratory analysis reveals however that the individual experience varies a great deal from the overall average. Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages. The wages data was used to illustrate exploratory longitudinal data analysis in Ozlem Ilk's PhD thesis supervised by author Cook (Ilk 2004), and was further developed into a case study for use in the teaching of exploratory data analysis at Iowa State University.

@@

This disparity between the average and the individual is a part of statistics, as a discipline, that is given less attention. It is this disparity which makes this particular data set important as a textbook example. Textbook data sets have longevity if there is a unresolved mystery. The iris data (Andereson 1935) is a prime example. It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis. (A side note: the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new and does not suffer from a connection with eugenics (Stodel 2020).) We argue that the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?

@@

As a textbook data set though, the wages data is outdated. The most recent year in the data is 1994, 10 years prior to when Singer and Willett (2003) was published. Teachers of statistics need use contemporary data sets to show how techniques are relevant for today's students. Using tired old textbook data sets can imbue a misconception that the field is not current. The wages data is extracted from NSLY79, one of the best examples of open data, which is constantly being updated. It should be possible to continuously refresh the textbook data from the data repository. This paper describes our (non-glamorous) journey from open (Open Knowledge Foundation 2021) wild data to textbook data.

@@

In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (\texttt{gender}), race (\texttt{race}), age (\texttt{age\_1979}), highest grade completed (\texttt{hgc}), highest grade completed in terms of years, e.g.~9th grade = 9, 3rd year college = 15, (\texttt{hgc\_i}), and the whether the graduate equivalency diploma is obtained (\texttt{ged}).

@@

The next step is tidying to obtain \texttt{hgc} and \texttt{ged}. As the highest grade completed is already exist in \texttt{new\_data\_qnames} and stored in years unit, we only rename the column name to be \texttt{hgc\_i} and recode it so we have the factor data type. For \texttt{ged}, we only separate the years and the GED status of the individual. Later on, this variable is useful to determine the refreshed high school dropouts data.

@@

@@

@@

\leavevmode\hypertarget{ref-nlsy79guide}{}%
---------. 2021b. ``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).'' Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University. Columbus, OH. \url{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}.

@@

>>>>

Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques. An example of this, is the wages data made public by Singer and Willett (2003) which can be used to teach generalized linear models, in addition to hierarchical, mixed effects and multilevel. The data records hourly wages of a sample of high school dropouts, from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S. Department of Labor 2021a).

The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average. An exploratory analysis reveals however that the individual experience varies a great deal from the overall average. Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages. The wages data was used to illustrate exploratory longitudinal data analysis in Ozlem Ilk's PhD thesis (Ilk 2004), and was further developed into a case study for use in the teaching of exploratory data analysis at Iowa State University.

@@

This disparity between the average and the individual is a part of statistics, as a discipline, that require more attention. It is this disparity which makes this particular data set important as a textbook example. Textbook data sets have longevity if there is a unresolved mystery. The iris data (Andereson 1935) is a prime example. It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis. (A side note: the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new and does not suffer from a connection with eugenics (Stodel 2020).) We argue that the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?

@@

As a textbook data set though, the wages data is outdated. The most recent year in the data is 1994, 10 years prior to when Singer and Willett (2003) was published. Teachers of statistics need contemporary data sets to show how techniques are relevant for today's students. Using tired old textbook data sets can imbue a misconception that the field is not current. The wages data is extracted from NSLY79, one of the best examples of open data, which is constantly being updated. It should be possible to continuously refresh the textbook data from the data repository. This paper describes our (non-glamorous) journey from open (Open Knowledge Foundation 2021) wild data to textbook data.

@@

In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (\texttt{gender}), race (\texttt{race}), age (\texttt{age\_1979}), highest grade completed (\texttt{hgc}), highest grade completed in terms of years, e.g.~9th grade = 9, 3rd year college = 15, (\texttt{hgc\_i}), and whether the graduate equivalency diploma is obtained (\texttt{ged}).

@@

The next step is tidying to obtain \texttt{hgc} and \texttt{ged}. As the highest grade completed is already exist in \texttt{new\_data\_qnames} and stored in year unit, we only rename the column name to be \texttt{hgc\_i} and recode it so we have the factor data type. For \texttt{ged}, we only separate the years and the GED status of the individual. Later on, this variable is useful to determine the refreshed high school dropouts data.

@@

In the original data, there is no explicit explanation of how the dropouts cohort is obtained from the database. Hence, we use the high school dropouts criteria mentioned in Wolpin (2005) to determine who is the dropouts, which are:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The individual whose highest grade completed (\texttt{hgc}) is reported to be less than 12th grade, \textbf{or}
\item
  The individual whose highest grade completed (\texttt{hgc}) is reported to be at least 12th grade and have received a GED (\texttt{ged} is code to 2).
\end{enumerate}

Further, Singer and Willett (2003) set additional criteria for the dropouts in the textbook data, which are male and aged between 14 and 17 years old in 1979. Hence, we also use this information to re-create the dropouts data.

With this filtering, we obtained 670 individuals in the refreshed data compared to 888 individuals in the original data. We then matched the individuals in the original and refreshed dataset to investigate why some individuals in the original data were not captured in the refreshed data. We found several reasons for this as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  173 individuals were more than 17 years old in 1979. Thus, it looks like the description of the original data is not quite accurate, that there are people older than 17 in their subset. Our decision is to also include them in the refreshed data as the data has age variable in it, so analyst could filter them later.
\item
  79 individuals were less than or equal to 17 years old in 1979. However, they were not captured in the refreshed data because:
\end{enumerate}

\begin{itemize}
\tightlist
\item
  35 of them completed at least 12th grade and having a diploma instead of GED (\texttt{ged} variable is coded to 1). Thus, they are indeed are not dropouts and we decide to not include them in the refreshed data.
\item
  The information of \texttt{ged} is missing in 38 individuals. We decide to include them in the refreshed data.
\item
  3 individuals are having both diploma and GED (\texttt{ged} is coded to 3). They essentially have GED, so we decide to keep them in the refreshed data as well.
\item
  12 individuals do not exist in \texttt{wages} data because they were participated in less than 3 rounds of survey.
\end{itemize}

We then redo the filtering based on those decisions we made on those individuals who were not captured in the refreshed data. Finally, we have 863 dropouts in the refreshed data, the closest number of individuals we get to the original data.

@@

\hypertarget{the-takeaways}{%
\subsection{The takeaways}\label{the-takeaways}}

@@

\leavevmode\hypertarget{ref-nlsy79guide}{}%
---------. 2021b. ``National Longitudinal Survey of Youth 1979 Cohort, Topical Guide to the Data.'' \url{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}.

@@

\leavevmode\hypertarget{ref-nlsy79edu}{}%
Wolpin, Kenneth I. 2005. ``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).'' Published by Bureau of Labor Statistics, U.S. Department of Labor. \url{https://www.bls.gov/opub/mlr/2005/02/art3full.pdf}.
